---
title: "Economic Data - Lecture 7"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```







## After this lecture you should be able to

1. Explain the difference between register and survey (or ILO) unemployment.

2. Explain how the unemployment rate is calculated.

3. Explain the intuition behind seasonal adjustment. 






---
class: inverse, center, middle

# Data About the Labor Market




---

## From population to labour force

The overall population of the UK ($\approx$ 66 mio in the UK)

- Children and elderly ($\approx$ 24 mio in the UK).

- Out of the labour force ($\approx$ 9 mio in the UK).

- The labour force ($\approx$ 33 mio in the UK) (also called the economic active).




---
## Out of the labour force

Also known as: The economic inactive

- Studying (≈ 27% in the UK)

- Looking after the family/staying home (≈ 24% in the UK) 

- Long-term sick (≈ 23% in the UK)

- Retired (≈ 13% in the UK)






---
## The labour force

**The labour force = Unemployed + Employed**

*Unemployed* (ILO definition) 

- Without a job, have been actively seeking work in the past four weeks and are available to start work in the next two weeks.

- Out of work, have found a job and are waiting to start it in the next two weeks.

*Employed* (ILO definition)

- Work at least one hour for pay in a week.

- Temporarily away from jobs.

- On Government-supported training/employment programs.

- Unpaid family work.


---

class: inverse, center, middle

$$\Large \text{Unemployment rate} = \frac{\text{Unemployed}}{\text{Unemployed} + \text{Employed}}$$


---


```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}


df <- read_csv('../data/AURUKM.csv') 

p1 <- ggplot(data = df, aes(x = DATE, y = AURUKM))+
  geom_line(lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "120 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Monthly unemployment rate for the UK, 1855-2016",
       caption = "Source: Bank of England, Unemployment Rate in the United Kingdom [AURUKM], retrieved from FRED, Federal Reserve Bank of St. Louis;\nhttps://fred.stlouisfed.org/series/AURUKM,November 10, 2023.")


```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p1
```







---


```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

df <- read_csv('../data/STLABOUR_10112023115611842.csv') %>% 
  select(TIME, Measure, Value) %>% 
  rename(date = TIME, 
         adj = Measure,
         unemp = Value) %>% 
  mutate(adj = ifelse(adj == 'Level, rate or quantity series', 'No seasonal adjustment', 'Seasonal adjustment'),
         date = paste(date, "01", sep="-"),
         date = as.Date(date, format = "%Y-%m-%d")) 

p1 <- ggplot(data = df %>% filter(adj == 'No seasonal adjustment'), aes(x = date, y = unemp))+
  geom_line(stat='identity',lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Monthly unemployment rate, Jan. 2000 - Jun. 2023",
       caption = "Source: OECD, Short-term Unemployment Statistics")
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p1
```




---


## Seasonal variation 

Labor market data has a substantial seasonal component.

- Night: fewer work, fewer travel etc.

- Weekends: fewer work, fewer commute, more in museums and at the beach.

- August: fewer work, more travel.

- December: private consumption is higher in UK.

This is an issue if we want to compare changes between periods. 

- Unemployment is always lower in December, does that mean "the economy" is doing better?

- We might want to understand if unemployment is lower *over and beyond* what we can expect from seasonal variation or the trend. 





---



```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

df <- read_csv('../data/STLABOUR_10112023115611842.csv') %>% 
  select(TIME, Measure, Value) %>% 
  rename(date = TIME, 
         adj = Measure,
         unemp = Value) %>% 
  mutate(adj = ifelse(adj == 'Level, rate or quantity series', 'No seasonal adjustment', 'Seasonal adjustment'),
         date = paste(date, "01", sep="-"),
         date = as.Date(date, format = "%Y-%m-%d")) 

p1 <- ggplot(data = df %>% filter(adj == 'No seasonal adjustment'), aes(x = date, y = unemp))+
  geom_line(stat='identity',lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Monthly unemployment rate, Jan. 2000 - Jun. 2023",
       subtitle = "No seasonal adjustment",
       caption = "Source: OECD, Short-term Unemployment Statistics")
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p1
```





---



```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

p1 <- ggplot(data = df %>% filter(adj == 'Seasonal adjustment'), aes(x = date, y = unemp))+
  geom_line(stat='identity',lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Monthly unemployment rate, Jan. 2000 - Jun. 2023",
       subtitle = "Seasonal adjustment",
       caption = "Source: OECD, Short-term Unemployment Statistics")
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p1
```




---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

p1 <- ggplot(data = df, aes(x = date, y = unemp, color = adj))+
  geom_line(stat='identity',lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Monthly unemployment rate, Jan. 2000 - Jun. 2023",
       caption = "Source: OECD, Short-term Unemployment Statistics")
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p1
```




---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

df1 <- df %>% 
  filter(adj == 'Seasonal adjustment') %>%
  select(unemp, date) %>% 
  rename(unemp_s = unemp) %>% 
  full_join(df %>% filter(adj != 'Seasonal adjustment'), by = 'date') %>% 
  mutate(diff = unemp - unemp_s) %>% 
  filter(date > '2016-01-01')

p1 <- ggplot(data = df1, aes(x = date, y = diff))+
  geom_line(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "4 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Monthly unemployment rate, Jan. 2016 - Jun. 2023",
       caption = "Source: OECD, Short-term Unemployment Statistics")


```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p1
```





---
## Decomposing a time series


We can decompose the variation in the unemployment rate $u_t$ as follows

$$u_t = s_t+\lambda_t +\varepsilon_t$$
- $\lambda_t$ is a trend component.

- $s_t$ is a seasonal compononent. 

- $\varepsilon_t$ are random shocks.

Some statistical agencies provide data where the different compononents have been isolated.

There is a wide nariety of ways to decompose a time series and the correct method depends on the series and application.

Let's create our own decomposition to get a sense of how it works. 



---

## Finding the trend component

One way to remove seasonality and other shocks is through the use of linear filters. 

One such filter is the moving average or MA filter.

The MA filter of order k or $MA(k)$ is given by

$$\lambda_t = \frac{1}{2k + 1}\sum_{t-k}^{t+k}y_t$$

A moving average is commonly used with time series data to smooth out short-term fluctuations and highlight longer-term trends.

Let's try this with our unemployment data.



---

```{r, include = F}
# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(lubridate)
library(ggpubr)


# Loading and cleaning the data 
df <- read_csv('../data/STLABOUR_10112023115611842.csv') %>% 
  select(TIME, Measure, Value) %>% 
  rename(date  = TIME, 
         adj   = Measure,
         unemp = Value) %>% 
  mutate(adj  = ifelse(adj == 'Level, rate or quantity series', 'No seasonal adjustment', 'Seasonal adjustment'),
         date = paste(date, "01", sep="-"),
         date = as.Date(date, format = "%Y-%m-%d")) %>% 
  mutate(month = month(date), year = year(date)) %>% 
  filter(adj == 'No seasonal adjustment') 


# Calcualting the trend 
df <- df %>% 
  mutate(trend = (lag(unemp, n=4)   + lag(unemp, n=3)   + lag(unemp, n=2)  + lag(unemp, n=1)  + 
                  lead(unemp, n=4)  + lead(unemp, n=3)  + lead(unemp, n=2) + lead(unemp, n=1) + 
                  unemp)/9) 

df <- df %>% 
  mutate(month = month(date), year = year(date)) %>% 
  group_by(month) %>% 
  summarise(unemp_month = mean(unemp - trend, na.rm = T)) %>% 
  full_join(df, by = 'month') %>% 
  mutate(unemp_shock = unemp - (trend + unemp_month)) %>% 
  filter(date >= '2014-01-01', date< '2020-03-01') 

```


`r chunk_reveal("chunk_1")`

```{r chunk_1, include = FALSE}
# Loading packages
library(tidyverse)
library(lubridate)

# Loading and cleaning the data
read_csv('../data/STLABOUR_10112023115611842.csv') %>% 
  select(TIME, Measure, Value) %>% 
  rename(date  = TIME) %>%  
  rename(adj   = Measure) %>% 
  rename(unemp = Value) %>% 
  mutate(adj  = ifelse(adj == 'Level, rate or quantity series', 'No seasonal adjustment', 'Seasonal adjustment')) %>% 
  mutate(date = paste(date, "01", sep="-")) %>% 
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>% 
  mutate(month = month(date), year = year(date)) %>% 
  filter(adj == 'No seasonal adjustment') %>% 
  mutate(trend = (lag(unemp, n=4)   + lag(unemp, n=3)   + lag(unemp, n=2)  + lag(unemp, n=1)  + 
                  lead(unemp, n=4)  + lead(unemp, n=3)  + lead(unemp, n=2) + lead(unemp, n=1) + 
                  unemp)/9) 

```

---


`r chunk_reveal("chunk_2")`

```{r chunk_2, include = FALSE}
# Plotting the data
ggplot(data = df, aes(x = date, y = trend))+
  geom_line(stat='identity', lwd = 1)+
    scale_x_date(date_breaks = "4 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemp. rate (%)") +
  labs(title = "Trend component",
       caption = "")+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())
```






---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

trend_plot <- ggplot(data = df, aes(x = date, y = trend))+
  geom_line(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "4 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemp. rate (%)") +
  labs(title = "Trend component for the UK",
       caption = "")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
trend_plot
```


---

## Finding the season component

To extract the season component, we can average over each month. 

The part of the unemployment not explained by the trend is given by

$$u_t - \lambda_t$$

Let's calcualte the average of this value for each month of the year. 

This will be our seasonal (monthly) component.




---





`r chunk_reveal("chunk_3")`

```{r chunk_3, include = FALSE}
df %>% 
  mutate(month = month(date), year = year(date)) %>% 
  group_by(month) %>% 
  summarise(unemp_month = mean(unemp - trend, na.rm = T)) %>% 
  full_join(df, by = 'month') 

```


---



`r chunk_reveal("chunk_4")`

```{r chunk_4, include = FALSE}
ggplot(data = df, aes(x = date, y = unemp_month))+
  geom_line(stat='identity')+
  theme(plot.title = element_text(size=12),
        plot.subtitle = element_text(size=9),
        legend.title = element_blank(),
        legend.position="bottom",
        plot.caption = element_text(size=6, hjust = 0),
        axis.title = element_text(size = 11),
        strip.background = element_blank(),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        legend.key = element_rect(fill = NA),
        axis.text.x=element_text(angle=45, hjust=1),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))+
  scale_x_date(date_breaks = "12 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Seasonal component",
       caption = "")

```


---


```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
season_plot <- ggplot(data = df, aes(x = date, y = unemp_month))+
  geom_line(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "4 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemp. rate (%)") +
  labs(title = "Seasonal component for the UK",
       caption = "")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
season_plot
```



---







## Finding the shock component

Finally, the shock is just what is not explained by the seasonal variation or the trend, 

$$\varepsilon = u_t - \lambda_t - s_t$$
This can be thought of as "surprises", something we did not expect by looking at the trend or the month of the year.

- This can be some positive or negative economic event for example.

This component can tell us if the labor market is doing better than what is explained by seasonal variation or just a trend.


---


`r chunk_reveal("chunk_5")`

```{r chunk_5, include = FALSE}
df %>% 
  mutate(unemp_shock = unemp - (trend + unemp_month)) 
```


---
`r chunk_reveal("chunk_6")`

```{r chunk_6, include = FALSE}
ggplot(data = df, aes(x = date, y = unemp_shock))+
  geom_line(stat='identity')+
  theme(plot.title = element_text(size=12),
        plot.subtitle = element_text(size=9),
        legend.title = element_blank(),
        legend.position="bottom",
        plot.caption = element_text(size=6, hjust = 0),
        axis.title = element_text(size = 11),
        strip.background = element_blank(),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        legend.key = element_rect(fill = NA),
        axis.text.x=element_text(angle=45, hjust=1),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))+
  scale_x_date(date_breaks = "12 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Shock component",
       caption = "Source: OECD, Short-term Unemployment Statistics")
```


---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

shock_plot <- ggplot(data = df, aes(x = date, y = unemp_shock))+
  geom_line(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "4 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemp. rate (%)") +
  labs(title = "Shock component for the UK",
       caption = "Source: OECD, Short-term Unemployment Statistics.")


```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
shock_plot
```


---

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
ggarrange(trend_plot, season_plot, shock_plot, ncol = 1, nrow = 3)
```





<!-- then show how to do that with the series for the uk, that way we get the trend  -->

<!-- then we remove the seasonality, just average by quarter and remove it  -->


---



## Seasonal adjustments: summary

Goal: remove seasonal component

Approach:

- Estimate trend level

- Remove trend from raw series

- What is left: seasonal component + error

- Use that to identify seasonal component

- Subtract the seasonal component from the raw series

There is a large and very advanced literature in econometrics and statistics on filtering time series!





---

## The labour market: other topics

Labour supply

- Extensive margin vs intensive margin

- Hours worked

Earnings

- Wages, Salary, survey based, register based

Gender disparities

- Earnings & promotions




---

## Summary

1. Explain the difference between register and survey (or ILO) unemployment.

2. Explain how the unemployment rate is calculated.

3. Explain the intuition behind seasonal adjustment. 








---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







