---
title: "Economic Data - Lecture 2"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is a recommended set up for flipbooks
knitr::opts_chunk$set(fig.width = 6, small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}

.centered-content {
  display: flex;
  flex-direction: column;
  justify-content: center;
  align-items: center;
  height: 60vh; /* Adjust based on your needs */
  text-align: center;
  margin-top: auto;
  margin-bottom: auto;
}

```


<div class="centered-content">
  <h2> </h2>
   <p>  Lab 1 <p> 

  <p>  Data on people<p> 

   <p> Fertility <p>

  <p> Life expectancy <p>

  <p> Working with data on people <p>
</div>







---

class: inverse, center, middle

# Lab 1 - Extension



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv')

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = gdp_pc))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv')

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = log(gdp_pc)))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

# Growth rates and logarithms

Why did we transform the time series of GDP per capita with the natural log? 

Let $y_{t}$ denotes GDP per capita in year $t$, then 
\begin{equation}
Y_{2018} = Y_{2017} (1+g),
\end{equation}
where $g$ denotes the growth rate. Then 
\begin{equation}
\frac{Y_{2018}}{Y_{2017}} =  1+g,
\end{equation}
Taking the natural log gives 
\begin{equation}
\text{log}(Y_{2018}) - \text{log}(Y_{2017}) = \text{log}(1+g)\approx g.
\end{equation}
Why is that? 

Remember that: $f(x+x_0) \approx f(x_0) + f^{\prime}(x_0)(x-x_0)$, $\text{log}(1)$, and $\partial \text{log}(x)/\partial x=1/x$.



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv') %>%
  mutate(d_l_gdp_pc = log(gdp_pc) - dplyr::lag(log(gdp_pc), n = 1),
         g_gdp_pc = (gdp_pc - dplyr::lag(gdp_pc, n = 1)) / dplyr::lag(gdp_pc, n = 1)) 

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = d_l_gdp_pc))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")+
  ylim(-0.055, 0.055)
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv') %>%
  mutate(d_l_gdp_pc = log(gdp_pc) - dplyr::lag(log(gdp_pc), n = 1),
         g_gdp_pc = (gdp_pc - dplyr::lag(gdp_pc, n = 1)) / dplyr::lag(gdp_pc, n = 1)) %>%
  pivot_longer(cols = c(gdp_pc, d_l_gdp_pc, g_gdp_pc), 
               names_to = "series", 
               values_to = "values") %>% 
  filter(!is.na(values))

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = values, linetype = series))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")+
  ylim(-0.055, 0.055)
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```


<!-- lab 1 plot, log, why logs? slope approximates the growth rate -->

<!-- why is that?  -->

<!-- consider GDP = ...  -->

<!-- very small approximately  -->

<!-- then plot the change, gives the growth rate,  -->

<!-- then plot the actual growth rate  -->

<!-- as long as g is small it works well, can see the correlation between the two variables  -->





---

class: inverse, center, middle

# Data About People



---

## What we use data about people for?

- Interesting in itself

- Used in the hockey stick chart (GDP per person!)

- Productivity measures

- Important for economic policy

- Pension schemes

- Tax base

- Budget/Expenditures (age composition)


Kowledge about demographic data is therefore important for many economic problems. 







---

# Census Data

- Census: We have to ask everyone.

- Super expensive and time-consuming (last census in the UK ~ £500 million)

- Only every 10 years in the UK since 1801

- Some countries have stopped running censuses (Last census in Denmark was in the 1970s)

ADD EHRE 



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
pop <- read_csv('../data/HISTPOP_27102023144550354.csv') %>% 
  rename(year = Time, pop = Value) %>% 
  dplyr::select(year, pop) %>% 
  mutate(pop = pop/1000000)

births <- read_xlsx('../data/annualreferencetables2021.xlsx', sheet = 6, skip = 5) %>% 
  rename(year = Year, 
         births = `Number of live births: United Kingdom`) %>% 
  mutate(year = as.numeric(year),
         Births = as.numeric(births)) %>% 
  select(year, Births) %>% 
  filter(year>=1980)

deaths <- read_xlsx('../data/annualreferencetables2021.xlsx', sheet = 7, skip = 5) %>% 
  rename(year = Year, 
         deaths = `Number of deaths: United Kingdom`) %>% 
  mutate(year = as.numeric(year),
         Deaths = as.numeric(deaths)) %>% 
  select(year, Deaths) %>% 
  filter(year>=1980)


# Combining datasets
vital_stats_wide <- births %>% 
  full_join(deaths, by = 'year')

vital_stats_long <- vital_stats_wide %>% 
  gather(variable, value, -year)


# Constructing plots 
p1 <- ggplot(data=pop, aes(x = year, y = pop))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("Population (millions)") +
  labs(title = "Total population of the UK, 1980-2021",
       subtitle = "Population is measured in millions",
       caption = "Source: OECD.")
  

p2 <- ggplot(data=vital_stats_wide, aes(x = year, y = Births-Deaths))+
  geom_line(size = 1.1, alpha = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Difference")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Vital statistics in the UK: births, deaths and marriages, Office of National Statistics (2023)")


p3 <- ggplot(data=vital_stats_long, aes(x = year, y = value, color = variable))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  xlab(" ") +
  ylab("") +
  labs(title = "Births and deaths in the UK, 1980-2021",
       subtitle = "",
       caption = "")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```


---


# Example: UK

**How many people were there in the UK on January 1, 2019?**
- Go to Eurostat and find: 66,647,112.


How do we know? Last census in the UK was in 2011...


---

# Components of change

We can decompose the change in poulation between $t$ and $t+\Delta t$ as follows: <br><br> <br><br>

\begin{equation}
\Delta \text{Pop}_{t+\Delta t, t} = (\underbrace{\text{births}_{t+\Delta t, t} - \text{deaths}_{t+\Delta t, t}}_{\text{Natural change}})+(\underbrace{\text{im}_{t+\Delta t, t} -  \text{em}_{t+\Delta t, t}}_{\text{Net migration}}) + \text{adj}_{t+\Delta t, t}.
\end{equation}
<br><br>
- $\text{adj}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births).

- $\text{births}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births).

<!-- - Embassies -->
<!-- - Military service -->
<!-- - International organizations -->
<!-- - Non-registered births -->






---

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
cowplot::plot_grid(p3, p2, nrow = 2, rel_heights = c(0.58, 0.42), align = "v")
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
migration <- read_excel("../data/december23ltimaccessible.xlsx", 
                        sheet = "Table 1", skip = 4) %>% 
  dplyr::select(Flow, Period, `All Nationalities`) %>% 
  rename(flow = `All Nationalities`) %>% 
  mutate(date = dmy(paste0("01-", gsub("YE ", "", Period)))) %>% 
  filter(Flow!='Net')



p1 <- ggplot(data=migration, aes(x = date, y = flow, color = Flow))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y") +  
  xlab(" ") +
  ylab("") +
  labs(title = "Migration and emigration in the UK, 2012-2023",
       subtitle = "",
       caption = "")


migration <- read_excel("../data/december23ltimaccessible.xlsx", 
                        sheet = "Table 1", skip = 4) %>% 
  dplyr::select(Flow, Period, `All Nationalities`) %>% 
  rename(flow = `All Nationalities`) %>% 
  mutate(date = dmy(paste0("01-", gsub("YE ", "", Period)))) %>% 
  filter(Flow=='Net')



p2 <- ggplot(data=migration, aes(x = date, y = flow, color = Flow))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y") +  
  xlab(" ") +
  ylab("Difference") +
  labs(title = "",
       subtitle = "",
       caption = "Source: Long-term international immigration, emigration and net migration flows, provisional, ONS (2024)")



```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
cowplot::plot_grid(p1, p2, nrow = 2, rel_heights = c(0.58, 0.42), align = "v")

```


---


class: inverse, center, middle

# Decomposing population changes

---


## Decomposing population changes

<br><br>

- Quantifying fertility trends <br><br>

- Quantifying mortality trends



---

## The crude birth rate

We can decompose the change in poulation between $t$ and $t+\Delta t$ as follows: <br><br>

\begin{equation}
\text{CBR}_{t} = \frac{\text{Births}_{t}}{\text{Pop.}_{t}}
\end{equation}

- $\text{adj}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births).

- $\text{births}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births).





---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
pop <- read_csv('../data/HISTPOP_27102023144550354.csv') %>% 
  rename(year = Time, pop = Value) %>% 
  dplyr::select(year, pop) 

n_births <- read_xlsx('../data/annualreferencetables2021.xlsx', sheet = 6, skip = 5) %>% 
  rename(year = Year, 
         births = `Number of live births: United Kingdom`) %>% 
  mutate(year = as.numeric(year),
         Births = as.numeric(births)) %>% 
  select(year, Births) %>% 
  filter(year>=1980) %>% 
  full_join(pop, by = 'year') 

crude_brate <-  n_births %>% 
  mutate(crude_brate = 1000*Births/pop)

p1 <- ggplot(data=crude_brate, aes(x = year, y = crude_brate))+
  geom_line(size = 1.1, alpha = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Births/female pop.")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Vital statistics in the UK: births, deaths and marriages, Office of National Statistics (2023)")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```

---




## The general fertility rate

\begin{equation}
\text{CBR}_{t} = \frac{\text{Births}_{t}}{\text{Pop.}_{t}}
\end{equation}

- $\text{adj}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births).

- $\text{births}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births).

Do we learn about behavior around fertility? 

consider .. 

but does not account for differences inside these brackets. 

also not suitable if we want to learn about changes in behavior vs. composition 

the question we are interested in matters! for waht data source we are interested in 




---

## Age specific fertility rate 

\begin{equation}
\text{AFR}_{g,t} = \frac{\text{Births}_{g,t}}{\text{Pop}_{gt}}
\end{equation}

expression 

to see how this matters consider the example 

could be that countries have similar fertility over the life cycle, but timing is different. 

- depends on marriage pattern, and various cultural factors. 

Might therefore still not be informative about behavior. 



---

<h2> </h2>
<table>
    <thead>
        <tr>
            <th>Age Group</th>
            <th>Number of Women (1000)</th>
            <th>Live Births (1000)</th>
            <th>Births per Woman</th>
        </tr>
    </thead>
    <tbody>
        <tr>
            <td colspan="4"><b>Baseline</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2000</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6000</b></td>
            <td><b>0.5000</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 1: Change in Age Composition</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>8000</td>
            <td>5333.333</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>4000</td>
            <td>1333.333</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6666.666</b></td>
            <td><b>0.5556</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 2: Change in Age-Specific Fertility Rate</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2500</td>
            <td>0.4167</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6500</b></td>
            <td><b>0.5417</b></td>
        </tr>
    </tbody>
</table>


---

## Total fertility rate 

A more common measure of a populations fertility is the total fertility rate: 
\begin{equation}
\text{TFR}_{t} = \sum_{g=1}^G\frac{\text{Births}_{g,t}}{\text{Pop}_{gt}}
\end{equation}
- Less sensitive to changes in the age and gender composition. 

- Therefore more informative about differences in fertility behavior.  <br><br>

$\Rightarrow$ Should I use the TFR, GFR, AFR, or TFR? Depends on the question. 







---

<h2> </h2>
<table>
    <thead>
        <tr>
            <th>Age Group</th>
            <th>Number of Women (1000)</th>
            <th>Live Births (1000)</th>
            <th>Births per Woman</th>
        </tr>
    </thead>
    <tbody>
        <tr>
            <td colspan="4"><b>Baseline</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2000</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6000</b></td>
            <td><b>0.5000</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 1: Change in Age Composition</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>8000</td>
            <td>5333.333</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>4000</td>
            <td>1333.333</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6666.666</b></td>
            <td><b>0.5556</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 2: Change in Age-Specific Fertility Rate</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2500</td>
            <td>0.4167</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6500</b></td>
            <td><b>0.5417</b></td>
        </tr>
    </tbody>
</table>





---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
tfr <- read_excel("../data/datadownload.xlsx",  skip = 6) %>% 
  rename(tfr = Rate, year = Year) %>% 
  filter(year>=1980)


p1 <- ggplot(data=tfr, aes(x = year, y = tfr))+
  geom_line(size = 1.1, alpha = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Total fertility rate")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Births in England and Wales from the Office for National Statistics (2022)")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---


## Measures of Mortality

Many of the same issues arise when we quantify mortality:

- Crude mortality rate: $\text{CMR}_t = \frac{\text{Deaths}_t}{\text{Pop}_t}$

- Age-specific mortality rate: $\text{CMR}_{xt} = \frac{\text{Deaths}_{xt}}{\text{Pop}_{xt}}$

- Infant mortality rate: $\text{IMR}_{t} = \frac{\text{Deaths}_{<1,t}}{\text{Pop}_{<1,t}}$ <br><br>

Depend on the age composition as well as the age specific mortality. 

Again, the best measure depends on the question at hand.  


---



## Life Expectancy

A very common measure of economic performance/development is life expectancy. 

Life expectancy measures average numbers of years left for people aged $x$:
\begin{equation}
e_x = \frac{T_x}{l_x}
\end{equation}
- where $T_x$ denotes the person-years left. 

- $l_x$ the number of people of age $x$.

Data for a range of years and countries are available.


---




## Life Expectancy

Two different types of ... 

How is life exepctancy calculated?

Need to get a brief overview of life tables to understand this.



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
life_exp <- read_csv("../data/Figure_1__Period_and_cohort_life_expectancy_at_birth,_males_and_females,_England_and_Wales,_1841_to_2020.csv", 
                     skip = 6) %>%  pivot_longer(cols = -Year, 
                                             names_to = "Category", 
                                             values_to = "Value") %>% 
  filter(Year>=1900)

p1 <- ggplot(data=life_exp, aes(x = Year, y = Value, color = Category))+
  geom_line(size = 1.1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Life expectancy")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Office for National Statistics – Expectation of life, England and Wales")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```




---






class: inverse, center, middle

# Summary


---

## Summary

You should know: 

- How this course is organized. 

- Why we need to learn about economic data. 

- Two principles of data visualization. 

- What R is. <br><br> 


**Next week**: Overview of demographic data. 





---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







