---
title: "Economic Data - Lecture 6"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is a recommended set up for flipbooks
knitr::opts_chunk$set(fig.width = 6, small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```




## Some announcements 

Have received some questions about the coursework.

The grade for the course will be based on the final coursework in December.

- The 20 quiz questions $10\%$

- 2 of the Mini Projects + 2 new Mini Projects $80\%$

- 1 feedback exercise. $10\%$

Released: November 27. Due: December 4.



---


## More About Data on Prices 

**Previous** lecture: 

- Price index: a weighted aggregation of many prices.

- Paasche and Laspeyres price index: considering the price of an initial basket of goods vs updating the basket.

- Chain-linked price indices.

- Nominal and real values: nominal does not account for changes in price levels, real does.


**This** lecture:

- Measuring prices over time.

- Measuring prices across space.




---
class: inverse, center, middle

# Measuring Prices Over Time




---

## Tracking Prices Over Time

We want to understand how prices evolve. Why?

- Calculate changes in volume for economic variables (keeping prices "fixed").

- Interesting in and of itself, for example to measure inflation. <br><br>


**Problem 1**: There are many prices in the economy. 

- Construct a price index weighted by expenditure shares. 

**Problem 2**: Both expenditure shares and prices change over time.

- Use the Paasche or Laspeyres indices. 

**Problem 3**: Expenditure shares fixed to a baseline become dated over time. 

- Use chain-linking. 


---


## Measuring Inflation  

All statistical agencies collect data on prices and inflation.

- The primary reason is for monetary and fiscal policy. 

The most common approach is the **annual inflation rate**:

$$\pi_t = \frac{P_{t} - P_{t-12}}{P_{t-12}},$$
where $P_{t}$ is some price index at time $t$.

The annual inflation rate compares prices for the latest month with the same month a year ago.

- We can download data for the UK from the August 2023 report [here](https://www.ons.gov.uk/economy/inflationandpriceindices/bulletins/consumerpriceinflation/august2023).



<!-- The most common approach to measuring inflation is the 12-month or annual inflation rate, which compares prices for the latest month with the same month a year ago -->




---


```{r, include = F}
# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)

df <- read_csv("../data/Figure_1__Annual_CPIH_and_CPI_inflation_rates_eased_slightly_in_August_2023.csv", skip =6) %>%
  rename(date    = `...1`) %>%
  mutate(date = paste(date, "01", sep=" "),
         date = as.Date(date, format = "%b %Y %d")) %>%
  gather(series, p, 2:4, -date)

p1 <- ggplot(data = df, aes(x = date, y = p, color = series))+
  geom_line(stat='identity', lwd = 1.2)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in UK",
       subtitle = "Measures monthly annual change in prices, Aug. 2013 - Aug. 2023.",
       caption = "Source: ONS.")
```


`r chunk_reveal("chunk_1")`

```{r chunk_1, include = FALSE}
# Loading packages
library(readxl)

# Loading and cleaning the data
df_inflation <- read.csv("../data/Figure_1__Annual_CPIH_and_CPI_inflation_rates_eased_slightly_in_August_2023.csv", skip = 6)
df_inflation %>% head()
names(df_inflation)[1] <- "date"
df_inflation %>% head()
df_inflation <- reshape(df_inflation,
                        varying   = names(df_inflation)[2:4],
                        v.names   = "p",             
                        timevar   = "series",         
                        times     = names(df_inflation)[2:4],  
                        direction = "long")
df_inflation %>% head()


```



---
`r chunk_reveal("chunk_2")`

```{r chunk_2, include = FALSE}
# Constructing a plot
ggplot(data = df, aes(x = date, y = p, color = series))+
  geom_line(stat='identity')+
    xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in UK")+
  labs(subtitle = "Measures monthly annual change in prices, Aug. 2013 - Aug. 2023.")+
  labs(caption = "Source: ONS.")+
  theme(plot.title = element_text(size=12))+
  theme(plot.subtitle = element_text(size=9))+
  theme(legend.title = element_blank())+
  theme(legend.position="bottom")+
  theme(plot.caption = element_text(size=6, hjust = 0))+
  theme(axis.title = element_text(size = 11))+
  theme(strip.background = element_blank())+
  theme(axis.text.x=element_text(angle=45, hjust=1))+
  theme(strip.placement = "outside")+
  theme(panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'))+
  theme(panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'))+
  theme(panel.background = element_rect(fill = "transparent", colour = NA))+
  theme(plot.background  = element_rect(fill = "transparent", colour = NA))+
  theme(legend.background = element_rect(fill='transparent'))+
  theme(legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")
```


---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning the main dataset
library(tidyverse)

df <- read_csv("../data/Figure_1__Annual_CPIH_and_CPI_inflation_rates_eased_slightly_in_August_2023.csv", skip =6) %>%
  rename(date    = `...1`) %>%
  mutate(date = paste(date, "01", sep=" "),
         date = as.Date(date, format = "%b %Y %d")) %>% 
  gather(series, p, 2:4, -date)  
  
p1 <- ggplot(data = df, aes(x = date, y = p, color = series))+
  geom_line(stat='identity', lwd = 1.2)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in UK",
       subtitle = "Measures monthly annual change in prices, Aug. 2013 - Aug. 2023.",
       caption = "Source: ONS.") 


```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning the main dataset
library(tidyverse)

df <- read_csv("../data/Figure_1__Annual_CPIH_and_CPI_inflation_rates_eased_slightly_in_August_2023.csv", skip =6) %>%
  rename(date    = `...1`) %>%
  mutate(date = paste(date, "01", sep=" "),
         date = as.Date(date, format = "%b %Y %d")) %>% 
  gather(series, p, 2:4, -date)  
  
p1 <- ggplot(data = df, aes(x = date, y = p, color = series))+
  geom_line(stat='identity', lwd = 1.2)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
  geom_hline(yintercept = 2, alpha = 0.5)+
  xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in UK",
       subtitle = "Measures monthly annual change in prices, Aug. 2013 - Aug. 2023.",
       caption = "Source: ONS.") 


```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```

---


### The UK Inflation Rate

The plot depicts a substantial increase in inflation over the last few years.

- **CPI**: Consumer Prices Index.

- **OOH**: Occupiers' housing costs.

- **CPIH**: CPI including owner occupiers' housing costs.

--

In all cases the increase is a weighted average of prices (Laspeyres).

- Weights are based on expenditure surveys and national account flows.

- We buy *more* of it $\Rightarrow$ it gets *higher* weight!

- Weights updated regularly and linked (chain-linking).








---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning the main dataset
df <- read_csv("../data/prc_hicp_midx__custom_8291826_linear.csv") %>%
  select(geo, TIME_PERIOD, OBS_VALUE, coicop) %>%
  rename(country = geo,
         date    = TIME_PERIOD,
         p       = OBS_VALUE,
         series  = coicop) %>%
  mutate(date = paste(date, "01", sep="-"),
         date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(country == 'FR') %>%
  group_by(series) %>%
  mutate(g_p = 100*(p-lag(p, n=12))/lag(p, n=12),
         g_p = round(g_p, d = 1))  %>%
  filter(date>'2018-01-01')



df_full <- read_csv("../data/prc_hicp_midx__custom_13484295_linear.csv") %>%
  select(geo, TIME_PERIOD, OBS_VALUE, coicop) %>%
  rename(country = geo,
         date    = TIME_PERIOD,
         p       = OBS_VALUE,
         series  = coicop) %>%
  mutate(date = paste(date, "01", sep="-"),
         date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(country == 'FR') %>%
  group_by(series) %>%
  mutate(g_p = 100*(p-lag(p, n=12))/lag(p, n=12),
         g_p = round(g_p, d = 1))  %>%
  filter(date>'2018-01-01')


p_full <- ggplot(data = df_full, aes(x = date, y = g_p, color = series))+
  geom_line(stat='identity', lwd = 1, show.legend = FALSE, alpha = 0.4)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in France by main aggregates",
       subtitle = "Measures monthly annual change in the HICP, Jan. 2018 - Oct. 2023.",
       caption = "Source: Eurostat, series prc_hicp_midx.")




p1 <- ggplot(data = df %>% filter(series == 'CP00'), aes(x = date, y = g_p))+
  geom_line(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in France",
       subtitle = "Measures monthly annual change in the HICP, Jan. 2018 - Oct. 2023.",
       caption = "Source: Eurostat, series prc_hicp_midx.")


p2 <- ggplot(data = df, aes(x = date, y = g_p, color = series))+
  geom_line(stat='identity', lwd = 1.2)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in France by main aggregates",
       subtitle = "Measures monthly annual change in the HICP, Jan. 2018 - Oct. 2023.",
       caption = "Source: Eurostat, series prc_hicp_midx.")+
  scale_color_discrete(labels=c('All items-HICP', 'Food', 'Energy', 'HICP ex. energy'))

```



`r chunk_reveal("chunk_3")`

```{r chunk_3, include = FALSE}
# Loading packages
library(tidyverse)
library(readxl)

# Loading and cleaning the data
read_csv("../data/prc_hicp_midx__custom_8291826_linear.csv") %>%
  select(geo, TIME_PERIOD, OBS_VALUE, coicop) %>%
  rename(country = geo) %>% 
  rename(date    = TIME_PERIOD) %>% 
  rename(p       = OBS_VALUE) %>% 
  rename(series  = coicop) %>%
  mutate(date = paste(date, "01", sep="-"),
         date = as.Date(date, format = "%Y-%m-%d")) %>%
  filter(country == 'FR') %>%
  group_by(series) %>%
  mutate(g_p = 100*(p-lag(p, n=12))/lag(p, n=12),
         g_p = round(g_p, d = 1))  %>%
  filter(date>'2018-01-01')
```



---
`r chunk_reveal("chunk_4")`

```{r chunk_4, include = FALSE}
# Constructing a plot
ggplot(data = df %>% filter(series == 'CP00'), aes(x = date, y = g_p))+
  geom_line(stat='identity', lwd = 1)+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Inflation rate (%)") +
  labs(title = "Monthly Inflation in France",
       subtitle = "Measures monthly annual change in the HICP, Jan. 2018 - Oct. 2023.",
       caption = "Source: Eurostat, series prc_hicp_midx.")+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())
```






---
```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```









---

## Some Challenges
   
Remember the general price index

$$P_{t} = \sum_{i=1}^N P_{it}Q_{it},$$
where $Q_{it}$ denotes the quantity and $P_{it}$ the price of good $i$.

--

**One** basket for everyone!

- But quantities are likely to be very different for different people.

- Policy response to inflation depends on the source of inflation.

- It is useful to consider more *disaggregated* indices to understand the source.

<!-- <br> -->




---
```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p_full
```






---
```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p2
```



---

## Summary

In all cases the increase is a weighted average of prices (Laspeyres).

- Weights are based on expenditure surveys and national account flows.

- We buy *more* of it $\Rightarrow$ it gets *higher* weight!

- Weights updated regularly and linked (chain-linking).

Then we study how the index changes over time.







---

class: inverse, center, middle

# Measuring Prices Across Space





---

## Measuring Prices Across Space

Like we for changes in prices to compare volumes over time, we can remove differences in prices across countries. 

- 100 euros can buy you less in Ireland than in Spain.

- Having double the money doesn’t make a difference if everything costs the double!


**Solution**: Compare the price of a representative basket of goods across locations.







---


```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}


df <- read_csv('../data/nama_10_pc__custom_8307407_linear.csv') %>% 
  select(geo, unit, OBS_VALUE) 


p1 <- ggplot(data = df %>% 
         filter(unit == 'CP_NAC_HAB') %>% 
         mutate(name = fct_reorder(geo, OBS_VALUE)), aes(x = name, y = OBS_VALUE))+
  geom_bar(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  xlab("") +
  ylab("GDP per capita") +
  labs(title = "GDP per capita 2022 (national currencies)",
       caption = "Source: Eurostat, series nama_10_pc")




p2 <- ggplot(data = df %>% 
               filter(unit == 'CP_EUR_HAB') %>% 
               mutate(name = fct_reorder(geo, OBS_VALUE)), aes(x = name, y = OBS_VALUE))+
  geom_bar(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  xlab("") +
  ylab("GDP per capita") +
  labs(title = "GDP per capita 2022 (euros)",
       caption = "Source: Eurostat, series nama_10_pc")





p3 <- ggplot(data = df %>% 
               filter(unit == 'CP_PPS_EU27_2020_HAB') %>% 
               mutate(name = fct_reorder(geo, OBS_VALUE)), aes(x = name, y = OBS_VALUE))+
  geom_bar(stat='identity', lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  xlab("") +
  ylab("GDP per capita") +
  labs(title = "GDP per capita 2022 (euros, PPS)",
       caption = "Source: Eurostat, series nama_10_pc")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---


```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p2
```

---



```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}

# Purchasing power parities 
df <- read_csv('../data/big-mac-full-index.csv') %>% 
  arrange(name, date) %>% 
  filter(!is.na(USD_adjusted)) %>% 
  mutate(USD_adjusted = 100*USD_adjusted,
         USD_raw = 100*USD_raw)



p_bm <- ggplot(df %>% 
               filter(date == '2022-07-01') %>% 
               mutate(name = fct_reorder(name, dollar_price)), 
             aes(x=name, y=dollar_price)) +
  # geom_hline(yintercept = 0) +
  geom_linerange(aes(ymin=0, ymax=dollar_price), alpha = 0.6, lwd = 1.1) +
  geom_point(alpha = 0.6, size=2.3) +
  theme_tufte() +
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("") +
  ylab("Current US Dollars") +
  labs(title = "The Big Mac Price",
       subtitle = "Dollar price of a Big Mac, selected currencies, %, July 2022",
       caption = "Source: The Economist")

```


```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p_bm
```

---

## Price Comparisons Across Regions

Consider a price index for a basket of goods in the home country

$$PPS = \sum_{i=1}^N P_iQ_i$$

Let the price of the basket in the foreign country be defined as 

$$PPS^* = \sum_{i=1}^N P_i^*Q_i^*$$

We assume that $Q_i^* = Q_i$ for all the goods $i$ $\Rightarrow$ compare the same basket in the two countries.

$PPS/PPS^*$ therefore gives the relative price of the basket in home versus foreign.



---


## Price Comparisons Across Regions

$PPS$ can be used to adjust prices for international comparison. 

$$\text{Price adjusted to PPS standards}_i = \frac{\text{Price national currency}_i}{PPS_i}$$

Can also be used to adjust prices to some base country

$$\text{Price adjusted to j country standards}_i = \frac{\text{Price national currency}_i}{PPS_i}\times PPS_{j}$$

A common application is to adjust nominal income differences across countries.



---

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---


```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p2
```



---


```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p3
```



---


```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}


if (!require(pacman)) install.packages("pacman")
p_load(eurostat,
       sf)


# Loading and cleaning GDP per capita data 
p <- read_csv('../data/prc_ppp_ind__custom_8308502_linear.csv') %>% 
  select(geo, TIME_PERIOD, OBS_VALUE) %>% 
  rename(CNTR_CODE = geo,
         year    = TIME_PERIOD, 
         p       = OBS_VALUE) 

gdpppc <- read_csv('../data/nama_10_pc__custom_8307407_linear.csv') %>% 
  select(geo, unit, OBS_VALUE) 


df_combined <- gdpppc %>% full_join(p %>% rename(geo = CNTR_CODE), by = 'geo') %>% 
  filter(unit == 'CP_PPS_EU27_2020_HAB')

SHP_0 <- get_eurostat_geospatial(resolution = 10, 
                                 nuts_level = 0, 
                                 year = 2013)


SHP_0_3035 <- st_transform(SHP_0, crs = 3035)


df <- SHP_0_3035 %>% full_join(p, by = 'CNTR_CODE')

map <- df %>% 
  ggplot(aes(fill = p)) +
  geom_sf(
    size = 0.05, 
    color = "white") +
  scale_fill_distiller(direction = 1,
                       guide = guide_colorbar(
                         direction = "horizontal",
                         title.position = "left",
                         barwidth = unit(4, "cm"),
                         barheight = unit(0.2, "cm"),
                         ticks = TRUE)) +
  scale_x_continuous(limits = c(2750000, 6500000)) +
  scale_y_continuous(limits = c(1530000, 5300000)) +
  theme_tufte() +
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.text=element_text(size=12),
        legend.position="bottom", 
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 11),
        strip.background = element_rect(fill="transparent", colour = NA),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        legend.key = element_rect(fill = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))+
  xlab("") +
  ylab("") +
  labs(title = "Price level index, 2022",
       # subtitle = "GDP is PPP adjusted at constant 2015 prices",
       fill = 'Price index (EU average = 100):',
       caption = "Source: Eurostat, series prc_ppp_ind")




p_comp <- ggplot(data = df_combined, 
       aes(x = OBS_VALUE, y = p))+  
    geom_point(alpha = 0.6, col = 'black', size=2, geom='point', stroke=0.2)+
    theme_tufte()+
    theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        # text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "6 month", date_labels = "%m-%Y")+
xlab("GDP per capita (PPS)") +
  ylab("Price index (EU average = 100)") +
  labs(title = "GDP per capita and Prices in Europe, 2022",
       caption = "Source: Eurostat, series prc_ppp_ind.")+
      scale_y_continuous(labels = scales::comma)+
        scale_x_continuous(labels = scales::comma)



```


```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
map
```



---



```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 14, fig.align="center"}
p_comp
```



---




## Price Comparisons Across Countries

International income differences tend to be smaller when adjusting for price differences.

This is known as the *Penn effect*.

- Comes from the *Penn World Tables* dataset (we worked with two weeks ago).

In economics, this is known as the Balassa–Samuelson effect

- The idea is that the difference between productivity in traded and non-traded sectors is higher in developed countries.

- Can show that this leads to higher prices in non-traded sectors in developed countries.

- Driven by traded and non-traded sectors competing for the same labor. 





---




<!-- is the proposition that this can be explained by the greater variation in productivity between developed and less developed countries in the traded goods' sectors which in turn affects wages and prices in the non-tradable goods sectors. -->



<!-- --- -->
<!-- ## PPP and exchange rates -->

<!-- We have seen one application of purchasing power parities so far: to make prices comparable across locations. -->




## Wide Data Formats

<div>
  <table>
    <caption>GDP by Country in Wide Format</caption>
    <thead>
      <tr>
        <th>Country</th>
        <th>GDP 2020</th>
        <th>GDP 2021</th>
        <th>GDP 2022</th>
      </tr>
    </thead>
    <tbody>
      <tr>
        <td>USA</td>
        <td>21000</td>
        <td>22000</td>
        <td>23000</td>
      </tr>
      <tr>
        <td>Canada</td>
        <td>1640</td>
        <td>1700</td>
        <td>1750</td>
      </tr>
      <tr>
        <td>Germany</td>
        <td>3850</td>
        <td>3900</td>
        <td>4000</td>
      </tr>
    </tbody>
  </table>
</div>

 <br><br>
 
- Displays GDP data across multiple years with each year in a separate column.

- Is often more readable and therefore provided when downloaded in Excel formats (like Eurostat).

- Preffered by statistical software: simplifies functions like grouping, filtering, and calculating trends over time.


---

## Long Data Formats

<div>
  <table>
    <caption>GDP by Country in Long Format</caption>
    <thead>
      <tr>
        <th>Country</th>
        <th>Year</th>
        <th>GDP</th>
      </tr>
    </thead>
    <tbody>
      <tr>
        <td>USA</td>
        <td>2020</td>
        <td>21000</td>
      </tr>
      <tr>
        <td>USA</td>
        <td>2021</td>
        <td>22000</td>
      </tr>
      <tr>
        <td>USA</td>
        <td>2022</td>
        <td>23000</td>
      </tr>
      <tr>
        <td>Canada</td>
        <td>2020</td>
        <td>1640</td>
      </tr>
      <tr>
        <td>Canada</td>
        <td>2021</td>
        <td>1700</td>
      </tr>
      <tr>
        <td>Canada</td>
        <td>2022</td>
        <td>1750</td>
      </tr>
      <tr>
        <td>Germany</td>
        <td>2020</td>
        <td>3850</td>
      </tr>
      <tr>
        <td>Germany</td>
        <td>2021</td>
        <td>3900</td>
      </tr>
      <tr>
        <td>Germany</td>
        <td>2022</td>
        <td>4000</td>
      </tr>
    </tbody>
  </table>
</div>






---


## Summary

After this lecture you should be able to:

- Know how the coursework works.

- How inflation is measured.

- Explain and use the PPS adjustment to compare prices across regions.

- Adjustment for price level differences across countries.

- Used to adjust monetary values across countries. 

We will work more on this material in this weeks lab.



---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







