---
title: "Economic Data - Lecture 9"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, fig.align="center",small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```

<style>

.center2 {
  margin: 0;
  position: absolute;
  top: 50%;
  left: 50%;
  -ms-transform: translate(-50%, -50%);
  transform: translate(-50%, -50%);
}

</style>



## After this lecture you should be able to

<br><br>

Explain the 3 different types of interest rates.

Explain why interest vary 

Describe exchange rates










---
class: inverse, center, middle

# Interest Rates 





---

## What is an interest rate?

<br><br>

Assume you borrow some amount from a bank. 

The interest rate is then 

$$ i = 100 \times \left(\frac{\text{amount to repay}}{\text{amount borrowed}} - 1\right)$$




---

## What is an interest rate?


Why do you have to pay interest to borrow? 

- Banks borrow and lend money on the money market. 

- Money in the future might be worth less due to inflation. 

- The bank could have invested the amount in stocks or bonds with positive interest.

- Risk of the loan not being repaid.



---


## Why are interest rates important?

Imagine you need **£10,000** to fund a business idea. <br><br>

Bank offer A:

- Payback **£10,050** in 10 years → **Interest rate: 0.5%**  
  
- $100 \times \left(\frac{10050}{10000} - 1 \right)$ <br><br>

Bank offer B:

- Payback **£20,000** in 10 years → **Interest rate: 100%**  
  
- $100 \times \left(\frac{20000}{10000} - 1 \right)$<br><br>


Need to be very certain about the profitability of the business idea in scenario B. 


  <!-- - Payback **£20,000** in 10 years → **Interest rate: 100%**   -->
  <!--   *Formula: \( 100\% = 100 \times \left( \frac{20,000}{10,000} - 1 \right) \)*   -->

---

## Impact on the Economy

- **Low Interest Rates**:

  - Cheaper loans → Stimulates business, car purchases, home renovations.<br><br>


- **High Interest Rates**:

  - Reduces borrowing → Slows economic activity.
  
  - Encourages savings.






---

## Annual or monthly interest rates

Data on interest rates are reported at different term periods.

Assume you borrow amount $B$ at time time and pay back $B^{\prime}$ in one year.

The annual interest: 

$$i^a = \frac{B^{\prime}}{B} - 1$$

The monthly interest: 

$$i^m = \left(\frac{B^{\prime}}{B}\right)^{\frac{1}{12}} - 1$$

The daily interest: 

$$i^d = \left(\frac{B^{\prime}}{B}\right)^{\frac{1}{365}} - 1$$








---

## Three important interest rates 

**Policy Rate**

- Set by the central bank

- Often based on an inflation target

- UK: "Bank of England Base Rate"<br><br>




**Interbank Interest Rate**

- Banks charge each other

- UK: "Sterling OverNight Indexed Average" or "SONIA"<br><br>



**Commercial lending rate**

- depends on repayment schedule, security/information, flexibility.






---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
library(lubridate)
library(tidyverse)
library(ggpubr)
library(ggthemes)

policy_rate <- read_csv("../data/IUMABEDR  Bank of England  Database.csv", 
                         col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Policy rate')



mortgage_rate <- read_csv("../data/IUMTLMV  Bank of England  Database.csv", 
                        col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. mortgage rate')


overdraft_rate <- read_csv("../data/IUMODTL  Bank of England  Database.csv", 
                          col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. overdraft rate')

rates <- rbind(policy_rate, mortgage_rate, overdraft_rate)

p1 <- ggplot(data = rates, aes(x = date, y = rate, linetype = type))+
  geom_line(lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, hjust = 0, angle = 45),
        axis.text.y=element_text(size = 14),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("interest rate (%)") +
  labs(title = "Interest rates in the United Kingdom, 1995-2023",
       caption = "Source: Bank of England. Series: IUMABEDR, IUMODTL and IUMTLMV.")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8.7, fig.width = 14, fig.align="center"}
p1
```





---
`r chunk_reveal("chunk_11")`

```{r chunk_11, include = FALSE}
read_csv("../data/IUMABEDR  Bank of England  Database.csv", 
                         col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Policy rate')



read_csv("../data/IUMTLMV  Bank of England  Database.csv", 
                        col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. mortgage rate')


overdraft_rate <- read_csv("../data/IUMODTL  Bank of England  Database.csv", 
                          col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. overdraft rate')

rates <- rbind(policy_rate, mortgage_rate, overdraft_rate)
```




---
`r chunk_reveal("chunk_12")`

```{r chunk_12, include = FALSE}
ggplot(data = rates, aes(x = date, y = rate, linetype = type))+
  geom_line(lwd = 1)+
    scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("interest rate (%)") +
  labs(title = "Interest rates in the United Kingdom, 1995-2023",
       caption = "Source: Bank of England. Series: IUMABEDR, IUMODTL and IUMTLMV.")+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, hjust = 0, angle = 45),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())
```





---

## Factors affecting the interest rate

There are many interest rates in the economy. 

There are many factors shaping the interest rates in the economy: 

- The cost of obtaining money (i.e. the base rate and the interbank interest rate).

- The risk of the loan.

- The time span of the loan.

- Whether the interest rate is fixed or flexible.







---

## The real interest rate

Data on interest rates are typically stated in nominal terms

- If I lend you money for a 2% annual interest rate and the inflation is 3% I lose.

- I can buy more for the 100£ I give you today, than for the 102£ you give me in one year.<br><br>



The link between the real interest rate $r$ and nominal interest $i$ rates are is the following: 

$$(1+i) = (1+r)(1+\pi) = 1 + r + \pi + \pi r$$

The *Fisher equation*: 

$$r = i - \pi$$

The indentity is based on the assumption that $\pi r\approx 0$.







---
class: inverse, center, middle

# Exchange Rates 

---



## Exchange rates

<br><br>

The rate at which we are willing to trade on currency to another currency

- “An exchange rate of 1.39 US Dollar to the British Pound.”

- “An exchange rate of 1.39 US Dollar to buy one British Pound.”




---





## Exchange rates

What determines the exchange rate?

- Financial institutions trade currencies on the foreign exchange market (forex).

- ... and then buy and sell them to us (we pay a premium)<br><br>

**Appreciates**: the currency increases in value – ex. 1£ from costing 1.39$ to 1.45$

**Depreciates**: the currency loses in value – ex. 1£ from costing 1.39$ to 1.28$<br><br>

Exchange rates, like many other prices, are determined by supply and demand.





---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
ex_rate <- read_table2("../data/Exchange_Rate_Report.tsv", 
                                    skip = 2) %>% 
  rename(rate = `U.K.`) %>% 
  mutate(date = as.Date(Date, format = "%d-%b-%Y")) %>% 
  select(date, rate) %>% 
  mutate(g = (rate - lag(rate, n = 1))/lag(rate, n = 1))


p1 <- ggplot(data = ex_rate, aes(x = date, y = rate))+
  geom_line()+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, angle=45, hjust=1),
        axis.text.y=element_text(size = 14),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("USD/GBP") +
  labs(title = "GBP against the US dollar, 1995-2023",
       caption = "Source: International Monetary Fund. Series: Representative rates.")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8.7, fig.width = 14, fig.align="center"}
p1
```









---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
ex_rate <- read_table2("../data/Exchange_Rate_Report.tsv", 
                                    skip = 2) %>% 
  rename(rate = `U.K.`) %>% 
  mutate(date = as.Date(Date, format = "%d-%b-%Y")) %>% 
  select(date, rate) %>% 
  mutate(g = (rate - lag(rate, n = 1))/lag(rate, n = 1))


p1 <- ggplot(data = ex_rate, aes(x = date, y = g))+
  geom_line(lwd = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, angle=45, hjust=1),
        axis.text.y=element_text(size = 14),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("USD/GBP") +
  labs(title = "GBP against the US dollar, 1995-2023",
       caption = "Source: International Monetary Fund. Series: Representative rates.")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8.7, fig.width = 14, fig.align="center"}
p1
```






---
`r chunk_reveal("chunk_21")`

```{r chunk_21, include = FALSE}
read_table2("../data/Exchange_Rate_Report.tsv", 
                                    skip = 2) %>% 
  rename(rate = `U.K.`) %>% 
  mutate(date = as.Date(Date, format = "%d-%b-%Y")) %>% 
  select(date, rate) %>% 
  mutate(g = (rate - lag(rate, n = 1))/lag(rate, n = 1))
```




---
`r chunk_reveal("chunk_22")`

```{r chunk_22, include = FALSE}
ggplot(data = ex_rate, aes(x = date, y = g))+
  geom_line(lwd = 0.5)+
    scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("USD/GBP") +
  labs(title = "GBP against the US dollar, 1995-2023",
       caption = "Source: International Monetary Fund. Series: Representative rates.")+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, angle=45, hjust=1),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())
```





---
## Exchange rates and interest rates

There is a link between interest rates and exchange rate: *uncovered interest rate parity*.

- It is essentially a non-arbitrage condition. 

For example a US investor can invest in the US at interest rate $i_t^{\text{us}}$. 

He could also invest in the UK at interest rate $i_t^{\text{uk}}$, but he needs to exchange at rate $E_t$.

*Uncovered interest rate parity* then states that 

\begin{equation}
1 + i_t^{\text{us}} = \frac{\mathbb{E}_t(E_{t+1})}{E_t}\left(1+i_t^{\text{uk}}\right),
\end{equation}
where $\mathbb{E}_t(E_{t+1})$ is the expected exchange rate in $t+1$ at time $t$.

$\Rightarrow$ Return from investing at home is the same of investing abroad in equilibrium. 





---
## Some date formats in R


%d = Day number of month (5, 17, 28, etc.)

%a = Abbreviated weekday (Mon, Tue, Wed, etc.)

%A = Full weekday (Monday, Tuesday, etc.) %w = Weekday number (0-6, Sunday is 0)

%u = Weekday number (1-7, Monday is 1)

%W = Week number (00-53, Monday is week start)

%U = Week number (01-53, Sunday is week start)

%m = Month number (e.g. 01, 02, 03, 04)

%b = Abbreviated month (Jan, Feb, etc.)

%B = Full month (January, February, etc.)

%y = 2-digit year (e.g. 89)

%Y = 4-digit year (e.g. 1989)




---
class: inverse, center, middle

# Summary





---



## Summary

Three important types of interest rates:

- Bank rate (Policy rate): The central bank’s rate.

- Interbank rate: the rate between banks.

- Household rate: the rate we are paying.


Why do interest rates vary?

- Information and security.

- Flexibility.

- Duration.


Describing exchange rates:

- Expressed in how much it costs to buy 1 currency unit.

- Appreciation/depcreciation: changes relative to other currencies.








---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







