---
title: "Economic Data - Lecture 2"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is a recommended set up for flipbooks
knitr::opts_chunk$set(fig.width = 6, small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}

.centered-content {
  display: flex;
  flex-direction: column;
  justify-content: center;
  align-items: center;
  height: 60vh; /* Adjust based on your needs */
  text-align: center;
  margin-top: auto;
  margin-bottom: auto;
}

```


<div class="centered-content">
  <h2> </h2>
   <p>  Lab 1 - Extension <p> 

  <p>  Data on people<p> 

   <p> Fertility <p>

  <p> Mortality <p>

  <p> Life expectancy <p>
  
   <p> Subsetting dataframes <p>

</div>







---

class: inverse, center, middle

# Lab 1 - Extension



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv')

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = gdp_pc))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv')

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = log(gdp_pc)))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

# Growth Rates and Logarithms

Why did we transform the time series of GDP per capita with the natural log? 

Let $Y_{t}$ denotes GDP per capita in year $t$, then 
\begin{equation}
Y_{2018} = Y_{2017} (1+g),
\end{equation}
where $g$ denotes the growth rate. Then 
\begin{equation}
\frac{Y_{2018}}{Y_{2017}} =  1+g,
\end{equation}
Taking the natural log gives 
\begin{equation}
\text{log}(Y_{2018}) - \text{log}(Y_{2017}) = \text{log}(1+g)\approx g.
\end{equation}
Why is $\text{log}(1+g)\approx g$? Because of the Taylor approximation. 

Remember that: $f(x) \approx f(x_0) + f^{\prime}(x_0)(x-x_0)$, $\text{log}(1)=0$, and $\partial \text{log}(x)/\partial x=1/x$.



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv') %>%
  mutate(d_l_gdp_pc = log(gdp_pc) - dplyr::lag(log(gdp_pc), n = 1),
         g_gdp_pc = (gdp_pc - dplyr::lag(gdp_pc, n = 1)) / dplyr::lag(gdp_pc, n = 1)) 

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = d_l_gdp_pc))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")+
  ylim(-0.055, 0.055)
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv') %>%
  mutate(d_l_gdp_pc = log(gdp_pc) - dplyr::lag(log(gdp_pc), n = 1),
         g_gdp_pc = (gdp_pc - dplyr::lag(gdp_pc, n = 1)) / dplyr::lag(gdp_pc, n = 1)) %>%
  pivot_longer(cols = c(gdp_pc, d_l_gdp_pc, g_gdp_pc), 
               names_to = "series", 
               values_to = "values") %>% 
  filter(!is.na(values))

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = values, linetype = series))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")+
  ylim(-0.055, 0.055)
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```





---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv') %>%
  filter(Year %in% c(1980, 1985, 1990, 1995, 2000, 2005, 2010, 2015)) %>% 
  mutate(d_l_gdp_pc = log(gdp_pc) - dplyr::lag(log(gdp_pc), n = 1),
         g_gdp_pc = (gdp_pc - dplyr::lag(gdp_pc, n = 1)) / dplyr::lag(gdp_pc, n = 1)) %>%
  pivot_longer(cols = c(d_l_gdp_pc, g_gdp_pc), 
               names_to = "series", 
               values_to = "values") %>% 
  filter(!is.na(values))

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = values, linetype = series))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita (log.)") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```


<!-- lab 1 plot, log, why logs? slope approximates the growth rate -->

<!-- why is that?  -->

<!-- consider GDP = ...  -->

<!-- very small approximately  -->

<!-- then plot the change, gives the growth rate,  -->

<!-- then plot the actual growth rate  -->

<!-- as long as g is small it works well, can see the correlation between the two variables  -->





---

class: inverse, center, middle

# Data About People



---

## Data About People


- Used to normalize data (e.g. GDP per capita)

- Productivity measures

- Important for economic policy

- Pension schemes

- Tax base

- Budget/Expenditures (age composition)<br><br>


Knowledge of and ability to work with demographic data is therefore important for understanding many economic issues. 







---

# Census Data

A census is a national enumeration of the population at the same time. 

- Census: We have to ask everyone.

- England and Wales since 1801 and the United States since 1790.

- By the 1980s around 94 percent of the world population was covered.

- Super expensive and time-consuming (last census in the UK ~ £500 million). Therefore only every decade.

- Some countries have stopped running censuses (The last census in Denmark was in the 1970s).




---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
pop <- read_csv('../data/HISTPOP_27102023144550354.csv') %>% 
  rename(year = Time, pop = Value) %>% 
  dplyr::select(year, pop) %>% 
  mutate(pop = pop/1000000)

births <- read_xlsx('../data/annualreferencetables2021.xlsx', sheet = 6, skip = 5) %>% 
  rename(year = Year, 
         births = `Number of live births: United Kingdom`) %>% 
  mutate(year = as.numeric(year),
         Births = as.numeric(births)) %>% 
  select(year, Births) %>% 
  filter(year>=1980)

deaths <- read_xlsx('../data/annualreferencetables2021.xlsx', sheet = 7, skip = 5) %>% 
  rename(year = Year, 
         deaths = `Number of deaths: United Kingdom`) %>% 
  mutate(year = as.numeric(year),
         Deaths = as.numeric(deaths)) %>% 
  select(year, Deaths) %>% 
  filter(year>=1980)


# Combining datasets
vital_stats_wide <- births %>% 
  full_join(deaths, by = 'year')

vital_stats_long <- vital_stats_wide %>% 
  gather(variable, value, -year)


# Constructing plots 
p1 <- ggplot(data=pop, aes(x = year, y = pop))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("Population (millions)") +
  labs(title = "Total population of the UK, 1980-2021",
       subtitle = "Population is measured in millions",
       caption = "Source: OECD.")
  

p2 <- ggplot(data=vital_stats_wide, aes(x = year, y = Births-Deaths))+
  geom_line(size = 1.1, alpha = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Difference")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Vital statistics in the UK: births, deaths and marriages, Office of National Statistics (2023)")


p3 <- ggplot(data=vital_stats_long, aes(x = year, y = value, color = variable))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  xlab(" ") +
  ylab("") +
  labs(title = "Births and deaths in the UK, 1980-2021",
       subtitle = "",
       caption = "")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```


---


# Example: UK

**How many people were there in the UK on January 1, 2019?**
- Go to Eurostat and find: 66,647,112.


How do we know? The last census in the UK was in 2011...


---

# Components of change

We can decompose the change in poulation between $t$ and $t+\Delta t$ as follows: <br><br> <br><br>

\begin{equation}
\Delta \text{Pop}_{t+\Delta t, t} = (\underbrace{\text{births}_{t+\Delta t, t} - \text{deaths}_{t+\Delta t, t}}_{\text{Natural change}})+(\underbrace{\text{im}_{t+\Delta t, t} -  \text{em}_{t+\Delta t, t}}_{\text{Net migration}}) + \text{adj}_{t+\Delta t, t}.
\end{equation}
<br><br>
- $\text{adj}_{t+\Delta t, t}$ denotes a statistical adjustment.

- The statistical adjustment accounts for staff in embassies, military service, non-registered births, and various other factors.

<!-- - $\text{births}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births). -->

<!-- - $\text{births}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births). -->

<!-- - $\text{births}_{t+\Delta t, t}$ denotes a statistical adjustment (embassies, military service, non-registered births). -->

<!-- - Embassies -->
<!-- - Military service -->
<!-- - International organizations -->
<!-- - Non-registered births -->






---

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
cowplot::plot_grid(p3, p2, nrow = 2, rel_heights = c(0.58, 0.42), align = "v")
```



---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
migration <- read_excel("../data/december23ltimaccessible.xlsx", 
                        sheet = "Table 1", skip = 4) %>% 
  dplyr::select(Flow, Period, `All Nationalities`) %>% 
  rename(flow = `All Nationalities`) %>% 
  mutate(date = dmy(paste0("01-", gsub("YE ", "", Period)))) %>% 
  filter(Flow!='Net')



p1 <- ggplot(data=migration, aes(x = date, y = flow, color = Flow))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y") +  
  xlab(" ") +
  ylab("") +
  labs(title = "Migration and emigration in the UK, 2012-2023",
       subtitle = "",
       caption = "")


migration <- read_excel("../data/december23ltimaccessible.xlsx", 
                        sheet = "Table 1", skip = 4) %>% 
  dplyr::select(Flow, Period, `All Nationalities`) %>% 
  rename(flow = `All Nationalities`) %>% 
  mutate(date = dmy(paste0("01-", gsub("YE ", "", Period)))) %>% 
  filter(Flow=='Net')



p2 <- ggplot(data=migration, aes(x = date, y = flow, color = Flow))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y") +  
  xlab(" ") +
  ylab("Difference") +
  labs(title = "",
       subtitle = "",
       caption = "Source: Long-term international immigration, emigration and net migration flows, provisional, ONS (2024)")



```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
cowplot::plot_grid(p1, p2, nrow = 2, rel_heights = c(0.58, 0.42), align = "v")

```


---


## Decomposing Population Changes

<br><br>

What are the underlying drivers of the population change we observe?

- This is crucial for a range of economic policies and questions in economics.<br><br>

We will therefore consider these issues in more depth:

- Quantifying fertility trends.

- Quantifying mortality trends.



---

## The Crude Birth Rate

The crude birth rate in year $t$ is denoted by: <br><br>

\begin{equation}
\text{CBR}_{t} = \frac{\text{Births}_{t}}{\text{Pop}_{t}}
\end{equation}

- $\text{Pop}_{t}$ denotes the total population in year $t$.

Notice, it does not account for: 

- the gender composition of the population or

- the age composition.

Impossible to say if there is a change in behavior or composition of the population using this measure.


---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
pop <- read_csv('../data/HISTPOP_27102023144550354.csv') %>% 
  rename(year = Time, pop = Value) %>% 
  dplyr::select(year, pop) 

n_births <- read_xlsx('../data/annualreferencetables2021.xlsx', sheet = 6, skip = 5) %>% 
  rename(year = Year, 
         births = `Number of live births: United Kingdom`) %>% 
  mutate(year = as.numeric(year),
         Births = as.numeric(births)) %>% 
  select(year, Births) %>% 
  filter(year>=1980) %>% 
  full_join(pop, by = 'year') 

crude_brate <-  n_births %>% 
  mutate(crude_brate = 1000*Births/pop)

p1 <- ggplot(data=crude_brate, aes(x = year, y = crude_brate))+
  geom_line(size = 1.1, alpha = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Births/Pop.")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Vital statistics in the UK: births, deaths and marriages, Office of National Statistics (2023)")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```

---




## The General Fertility Rate

The general fertility rate is given by <br><br>
\begin{equation}
\text{GFR}_{t} = \frac{\text{Births}_{t}}{\text{Female pop}_{15-44,t}}
\end{equation}

- $\text{Female pop}_{15-44,t}$ denotes the female population between $15$ and $44$ years of age.<br><br>

This is an improvement as it corrects for the gender composition. 

Ignores the fact that births are not equally likely between the ages $15$ and $44$.

The average age of giving birth in the UK in 2024 is $32$.

Variation in $\text{GFR}_{t}$ could therefore still be driven by age composition effects.






---

## Age-specific fertility rate 

The age-specific fertility rate is given by <br><br>

\begin{equation}
\text{AFR}_{g,t} = \frac{\text{Births}_{g,t}}{\text{Female pop}_{gt}}
\end{equation}

- where $g$ denotes the age group.  <br><br>

This is a useful measure as it can tell us where changes in births arise. 

Consider the following example.


<!-- to see how this matters consider the example  -->

<!-- could be that countries have similar fertility over the life cycle, but timing is different.  -->

<!-- - depends on marriage pattern, and various cultural factors.  -->

<!-- Might therefore still not be informative about behavior.  -->



---

<h2> </h2>
<table>
    <thead>
        <tr>
            <th>Age Group</th>
            <th>Number of Women </th>
            <th>Live Births </th>
            <th>Births per Woman</th>
        </tr>
    </thead>
    <tbody>
        <tr>
            <td colspan="4"><b>Baseline</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2000</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6000</b></td>
            <td><b>0.5000</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 1: Change in Age Composition</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>8000</td>
            <td>5333.333</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>4000</td>
            <td>1333.333</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6666.666</b></td>
            <td><b>0.5556</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 2: Change in Age-Specific Fertility Rate</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2500</td>
            <td>0.4167</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6500</b></td>
            <td><b>0.5417</b></td>
        </tr>
    </tbody>
</table>


---

## Total fertility rate 

A more common measure of a population fertility is the total fertility rate: 
\begin{equation}
\text{TFR}_{t} = \sum_{g=1}^G\frac{\text{Births}_{g,t}}{\text{Pop}_{gt}}
\end{equation}
- Less sensitive to changes in the age and gender composition. 

- Therefore more informative about differences in fertility behavior.  <br><br>

$\Rightarrow$ Should I use the TFR, GFR, AFR, or TFR? Depends on the question. 







---

<h2> </h2>
<table>
    <thead>
        <tr>
            <th>Age Group</th>
            <th>Number of Women (1000)</th>
            <th>Live Births (1000)</th>
            <th>Births per Woman</th>
        </tr>
    </thead>
    <tbody>
        <tr>
            <td colspan="4"><b>Baseline</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2000</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6000</b></td>
            <td><b>0.5000</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 1: Change in Age Composition</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>8000</td>
            <td>5333.333</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>4000</td>
            <td>1333.333</td>
            <td>0.3333</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6666.666</b></td>
            <td><b>0.5556</b></td>
        </tr>
        <tr>
            <td colspan="4"><b>Scenario 2: Change in Age-Specific Fertility Rate</b></td>
        </tr>
        <tr>
            <td>15-30</td>
            <td>6000</td>
            <td>4000</td>
            <td>0.6667</td>
        </tr>
        <tr>
            <td>31-45</td>
            <td>6000</td>
            <td>2500</td>
            <td>0.4167</td>
        </tr>
        <tr>
            <td><b>Total</b></td>
            <td><b>12000</b></td>
            <td><b>6500</b></td>
            <td><b>0.5417</b></td>
        </tr>
    </tbody>
</table>





---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
tfr <- read_excel("../data/datadownload.xlsx",  skip = 6) %>% 
  rename(tfr = Rate, year = Year) %>% 
  filter(year>=1980)


p1 <- ggplot(data=tfr, aes(x = year, y = tfr))+
  geom_line(size = 1.1, alpha = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Total fertility rate")+
  labs(title = "",
       subtitle = "",
       caption = "Source: Births in England and Wales from the Office for National Statistics (2022)")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```



---


## Measures of Mortality

Many of the same issues arise when we quantify mortality:

- Crude mortality rate: $\text{CMR}_t = \frac{\text{Deaths}_t}{\text{Pop}_t}$

- Age-specific mortality rate: $\text{CMR}_{xt} = \frac{\text{Deaths}_{xt}}{\text{Pop}_{xt}}$

- Infant mortality rate: $\text{IMR}_{t} = \frac{\text{Deaths}_{<1,t}}{\text{Pop}_{<1,t}}$ <br><br>

Depends on the age composition as well as the age-specific mortality. 

Again, the best measure depends on the question at hand.  


---



## Life Expectancy

A very common measure of economic performance/development is life expectancy. 

Life expectancy measures the average number of years left for people aged $x$:
\begin{equation}
e_x = \frac{T_x}{l_x}
\end{equation}
- where $T_x$ denotes the person-years left at age $x$ (sum of years left to live for the entire group). 

- $l_x$ the number of people of age $x$.

Data for a range of years and countries are available.


---




## Life Expectancy

Where do $T_x$ and $l_x$ come from? Life tables. 

- Tables that measure how the population diminishes, such as population number and age increases. <br><br>

Two different types of life tables: 

- **Period**: Fixes the age-specific mortality rates for one particular year.  

- **Cohort**: Uses the age-specific mortality rate of an actual cohort. <br><br>

Notice that cohort life tables contain predictions of future mortality rates for a cohort. 

Important to know which one to use, depending on the question. 

Cohort life expectancy takes economic development into account and is therefore higher. 





---

<br><br>

<h2> </h2>
<table>
    <thead>
        <tr>
            <th>Age (x)</th>
            <th>Number Alive at Start of Age (l<sub>x</sub>)</th>
            <th>Number Dying During Age Interval (d<sub>x</sub>)</th>
            <th>Probability of Dying (q<sub>x</sub>)</th>
            <th>Life Expectancy (e<sub>x</sub>)</th>
        </tr>
    </thead>
    <tbody>
        <tr>
            <td>0</td>
            <td>100,000</td>
            <td>1,000</td>
            <td>0.010</td>
            <td>75.0</td>
        </tr>
        <tr>
            <td>1</td>
            <td>99,000</td>
            <td>500</td>
            <td>0.005</td>
            <td>74.5</td>
        </tr>
        <tr>
            <td>2</td>
            <td>98,500</td>
            <td>100</td>
            <td>0.001</td>
            <td>73.6</td>
        </tr>
        <tr>
            <td>3</td>
            <td>98,400</td>
            <td>50</td>
            <td>0.0005</td>
            <td>72.6</td>
        </tr>
        <tr>
            <td>4</td>
            <td>98,350</td>
            <td>30</td>
            <td>0.0003</td>
            <td>71.6</td>
        </tr>
        <tr>
            <td>5</td>
            <td>98,320</td>
            <td>20</td>
            <td>0.0002</td>
            <td>70.6</td>
        </tr>
        <tr>
            <td>...</td>
            <td>...</td>
            <td>...</td>
            <td>...</td>
            <td>...</td>
        </tr>
    </tbody>
</table>


---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
life_exp <- read_csv("../data/Figure_1__Period_and_cohort_life_expectancy_at_birth,_males_and_females,_England_and_Wales,_1841_to_2020.csv", 
                     skip = 6) %>%  pivot_longer(cols = -Year, 
                                             names_to = "Category", 
                                             values_to = "Value") %>% 
  filter(Year>=1900)

p1 <- ggplot(data=life_exp, aes(x = Year, y = Value, color = Category))+
  geom_line(size = 1.1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_y_continuous(label=comma)+
  labs(x = "",
       y = "Life expectancy")+
  labs(title = "Period and cohort life expectancy in England and Wales, 1900-2020",
       subtitle = "",
       caption = "Source: Office for National Statistics – Expectation of life, England and Wales")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```




---






class: inverse, center, middle

# Back to R





---

## Summary

Understanding demographic data is key for a range of economic questions.

In this lecture, we have learned how to decompose changes in population.<br><br>

You should know: 

- Decomposing population growth. 

- Measuring trends in fertility. 

- Measuring trends in mortality. 

- Measuring life expectancy

- R: Subsetting dataframes.<br><br>


**Next week**: Measuring economic activity. 





---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







