---
title: "Economic Data - Lecture 3"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is a recommended set up for flipbooks
knitr::opts_chunk$set(fig.width = 6, small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}

.centered-content {
  display: flex;
  flex-direction: column;
  justify-content: center;
  align-items: center;
  height: 60vh; /* Adjust based on your needs */
  text-align: center;
  margin-top: auto;
  margin-bottom: auto;
}

```


<div class="centered-content">
  <h2> </h2>
   <p>  Lab 2 - Extension <p> 

  <p>  Measuring Economic Actvity<p> 

   <p> Fertility <p>

  <p> Mortality <p>

  <p> Life expectancy <p>
  
   <p> Subsetting dataframes <p>

</div>







---

class: inverse, center, middle

# Lab 1 - Extension


---

tips for writing scripts 





---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/economic_data_uk.csv')

# Constructing plots 
p1 <- ggplot(data=df, aes(x = Year, y = gdp_pc))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  xlab("Year") +
  ylab("GDP Per Capita") +
  labs(title = "GDP Per Capita UK, 1980-2018",
       caption = "Source: Maddison (2021).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```





---

class: inverse, center, middle

# Measuring Economic Activity



---

## Measuring Economic Activity


GDP, Gross Domestic Product, is the most frequently used indicator in the national accounts

- How does GDP measure economic activity?

- What are the main approaches to measuring GDP?

- How do we calculate the GDP of an economy?


---


## Total Value of Production

<!-- *GDP measures the value of all the output carried out by all the firms, non-profit, institutions, government bodies and households in a given country during a given period.* -->

A simple measure would be the sum of the value of of output for all firms:
\begin{equation}
\sum_{i = 1}^N Y_{it}P_{it}.
\end{equation}

- where $N$ is the number of firms, $Y_{it}$ production, and $P_{it}$ the price.

- $N$ are all the firms with the country.

- $t$ refers to a year or a quarter. <br><br>

This is not how GDP is defined though. Why?


---



## A Very Simple Economy

<table style="border-collapse: collapse; width: 100%; text-align: center; font-family: sans-serif;">
  <thead>
    <tr>
      <th colspan="3" style="border-bottom: 1px solid black;">Year 1</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td colspan="3">Firm A</td>
    </tr>
    <tr>
      <td>Output</td>
      <td> $ 100,000</td>
      <td></td>
    </tr>

    <tr>
      <td colspan="3" style="border-bottom: 1px solid black; padding: 10px;"></td>
    </tr>

    <tr>
      <th colspan="3" style="border-bottom: 1px solid black;">Year 2</th>
    </tr>
    <tr>
      <td></td>
      <td>Firm A1</td>
      <td>Firm A2</td>
    </tr>
    <tr>
      <td>Output</td>
      <td>$ 30,000</td>
      <td>$ 100,000</td>
    </tr>
  </tbody>
</table>

<br><br>
<!-- Firm A produces 100,000 worth of pasta in year 1.  -->

<!-- The firm splits into A1 and A2 in year 2 -->

<!-- - A1 produces 30,000 worth of flour. -->

<!-- - A2 buys it and sells total output for 100,000 -->


Splitting the firm increases GDP by $30$ percent!

<!-- The same quantity of pasta was produced at the same price. All that changed was the legal and commercial organisation of the firms. -->


---

## Measuring GDP

For this reason, we consider the sum of value added in the economy instead:
\begin{equation}
\text{GDP}_t = \sum_{i = 1}^N \text{Value added}_{it}
\end{equation}
measuring the value that the firm adds to that of the firms that supply its inputs.

Why does this work?

<!-- GDP combines in a single figure, and with no double counting, all the output (or production) carried out by all the firms, non-profit, institutions, government bodies and households in a given country during a given period,  -->

<!-- regardless of the type of goods and services produced, provided that the production takes place within the country’s economic territory. -->


---

## A Very Simple Economy

<table style="border-collapse: collapse; width: 100%; text-align: center; font-family: sans-serif;">
  <thead>
    <tr>
      <th colspan="3" style="border-bottom: 1px solid black;">Year 1</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td colspan="3">Firm A</td>
    </tr>
    <tr>
      <td>Value added</td>
      <td> $ 100,000</td>
      <td></td>
    </tr>

    <tr>
      <td colspan="3" style="border-bottom: 1px solid black; padding: 10px;"></td>
    </tr>

    <tr>
      <th colspan="3" style="border-bottom: 1px solid black;">Year 2</th>
    </tr>
    <tr>
      <td></td>
      <td>Firm A1</td>
      <td>Firm A2</td>
    </tr>
    <tr>
      <td>Value added</td>
      <td>$ 30,000</td>
      <td>$ 100,000 - 30,000 = 70,000</td>
    </tr>
  </tbody>
</table>

<br><br>

GDP remains unchanged when the organization of the economy changes. 



---



## The production approach

The production approach defines GDP in the following manner, 
\begin{equation}
\text{GDP}_t^P = \sum_{i = 1}^N \text{Value added}_{it} + NT
\end{equation}
- where $N$ denotes the set of production entities in the economy (firms etc.) 

- $NT$ denotes taxes minus subsidies. 

- Crucially it avoids double counting 

- Why $NT$? Taxes are subtracted from value added, need to adjust for this. 


---


## The Production Boundary

How about $N$, what counts as economic activity? The ESA 2010 standards has the answer:

- Does the activity produce a useful output?

- Is the product or activity marketable and does it have a market value?

- If the product does not have a meaningful market value, can one be assigned (imputed)?

- Would exclusion (or inclusion) of the product of the activity make comparisons between countries over time more meaningful?

<br><br> Not always obvious what is accounted for in GDP and what is not. 

- Housework, voluntary work, and criminal activities fall outside. 

<!-- Economic activity within the country. GNI captures activity in other countries.  -->

<!-- Does not account for depreciation (net domestic product).  -->


---

## Crime stuff here 






---



## The Expenditure approach

In the simple economy all expenditure equals value added $\Rightarrow$ GDP equals expenditure on pasta

In a real economy there are more ways to spend, this is the basis of the expenditure approach
\begin{equation}
\text{GDP}_t^E = C + I + G + Ex - Im
\end{equation}
- C the final consumption of goods and services by households

- is the final consumption expenditure of goods and services by the government

- I is investment (gross capital formation)

- Ex are exports and Im imports


---


## The Income approach

In the simple economy, income from pasta is the only income in the eocnomy.  

In a bigger economy, it can be written 
\begin{equation}
\text{GDP}_t^I =W + P + NT
\end{equation}
- W denotes worker compensation 

- P denotes profits

- NT taxes minus subsidies



---

## Reconciling output and income

\begin{equation}
\text{GDP}_t^P  = \text{GDP}_t^E = \text{GDP}_t^I =
\end{equation}






---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/nama_10_gdp__custom_13032981_linear.csv')

# Constructing plots 
p1 <- ggplot(data=df, aes(x = TIME_PERIOD, y = OBS_VALUE))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="none",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
    scale_y_continuous(label=comma)+
  xlab("Year") +
  ylab("Million GBP") +
  labs(title = "GDP UK in value (current prices), 2000-2019",
       caption = "Source: Eurostat (nama_10_gdp).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```


---

## GDP in Volume

Does the increase in GDP stem from changes in quantity or prices? 

GDP at current prices can be adjusted using the following identity,
\begin{equation}
\text{GDP deflator}_t = \frac{\text{Nominal GDP}_t }{\text{Real GDP}_t } 
\end{equation}
- Note this is different from the CPI, not all goods in GDP are in the CPI.

- Very important to know if GDP is in volume or value. 

- Economists typically care about real GDP. However, value is often used to standardize different variables. 

MORE ON THIS LATER


---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/nama_10_gdp__custom_13033988_linear.csv') %>% 
  mutate(unit = ifelse(unit == 'CP_MNAC', 'Current prices', unit),
         unit = ifelse(unit == 'CLV10_MNAC', '2010 prices', unit))

# Constructing plots 
p1 <- ggplot(data=df, aes(x = TIME_PERIOD, y = OBS_VALUE, linetype = unit))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="right",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
    scale_y_continuous(label=comma)+
  xlab("Year") +
  ylab("Million GBP") +
  labs(title = "GDP UK, 2000-2019",
       caption = "Source: Eurostat (nama_10_gdp).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```






---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
# Loading and cleaning data 
df <- read_csv('../data/nama_10_gdp__custom_13033988_linear.csv') %>% 
  pivot_wider(names_from = unit, values_from = OBS_VALUE) %>% 
  mutate(gdp_def = CP_MNAC/CLV10_MNAC)

# Constructing plots 
p1 <- ggplot(data=df, aes(x = TIME_PERIOD, y = gdp_def*100))+
  geom_line(size = 1.1, alpha = 0.7)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="right",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1, size = 12),
        axis.text.y=element_text(size = 12),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
    scale_y_continuous(label=comma)+
  xlab("Year") +
  ylab("GDP deflator (2010 = 100)") +
  labs(title = "GDP Deflator UK, 2000-2019",
       caption = "Source: Eurostat (nama_10_gdp).")
  
```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8, fig.width = 11, fig.align="center"}
p1
```




## What to use GDP for? 



---

## Why GDP?

Gross consumption of fixed capital=not dectued, that's net value added 

domestic, GNI etc. for example working somewhere else 

product, refers to production 



---






---

## Summary

Understanding demographic data is key for a range of economic questions.

In this lecture, we have learned how to decompose changes in population.<br><br>

You should know: 

- Decomposing population growth. 

- Measuring trends in fertility. 

- Measuring trends in mortality. 

- Measuring life expectancy

- R: Subsetting dataframes.<br><br>


**Next week**: Measuring economic activity. 





---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







