---
title: "Economic Data - Lecture 8"
subtitle: "Sebastian Ellingsen"
author: "School of Economics - University of Bristol"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: true # false if wish to disable scrolling by mouth and use keyboard only

---

```{r xaringan-themer, include=FALSE, warning=FALSE}
options(htmltools.dir.version = FALSE)

library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253")

options(htmltools.dir.version = FALSE)

# This is the recommended set up for flipbooks
# you might think about setting cache to TRUE as you gain practice --- building flipbooks from scratch can be time consuming
knitr::opts_chunk$set(fig.width = 6, fig.align="center",small.mar=TRUE, fig.retina = 5,message = FALSE, warning = FALSE, comment = "", cache = F)
library(flipbookr)
library(tidyverse)
library(readxl)
library(scales)
library(ggthemes)
```


```{css, eval = TRUE, echo = FALSE}
.remark-code{line-height: 1.5; font-size: 80%}

@media print {
  .has-continuation {
    display: block;
  }
}

code.r.hljs.remark-code{
  position: relative;
  overflow-x: hidden;
}


code.r.hljs.remark-code:hover{
  overflow-x:visible;
  width: 500px;
  border-style: solid;
}
```

<style>

.center2 {
  margin: 0;
  position: absolute;
  top: 50%;
  left: 50%;
  -ms-transform: translate(-50%, -50%);
  transform: translate(-50%, -50%);
}

</style>



## After this lecture you should be able to

Explain the 3 different types of interest rates.

Explain why interest vary 

Describe exchange rates










---
class: inverse, center, middle

# Interest Rates 





---

## What is an interest rate?

Assume you borrow some amount from a bank. 

The interest rate is then 

$$ i = \frac{\text{amount to repay}}{\text{amount borrowed}} - 1$$

Why do you have to pay interest to borrow? 

- Banks borrow and lend money on the money market. 

- Money in the future might be worth less due to inflation. 

- The bank could have invested the amount in stocks or bonds.

- Risk of the loan not being repaid.


---

## Annual or monthly interest rates

Data on interest rates are reported at different term periods.

Assume you borrow amount $B$ at time time and pay back $B^{\prime}$ in one year.

The annual interest: 

$$i^a = \frac{B^{\prime}}{B} - 1$$

The monthly interest: 

$$i^m = \left(\frac{B^{\prime}}{B}\right)^{\frac{1}{12}} - 1$$

The daily interest: 

$$i^d = \left(\frac{B^{\prime}}{B}\right)^{\frac{1}{365}} - 1$$








---

## Three important interest rates 

**Policy Rate**

- Set by the central bank

- Often based on an inflation target

- UK: "Bank of England Base Rate"



**Interbank Interest Rate**

- Banks charge each other

- UK: "Sterling OverNight Indexed Average" or "SONIA"


**Commercial lending rate**

- depends on repayment schedule, security/information, flexibility.






---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
library(lubridate)
library(tidyverse)
library(ggpubr)
library(ggthemes)

policy_rate <- read_csv("../data/IUMABEDR  Bank of England  Database.csv", 
                         col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Policy rate')



mortgage_rate <- read_csv("../data/IUMTLMV  Bank of England  Database.csv", 
                        col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. mortgage rate')


overdraft_rate <- read_csv("../data/IUMODTL  Bank of England  Database.csv", 
                          col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. overdraft rate')

rates <- rbind(policy_rate, mortgage_rate, overdraft_rate)

p1 <- ggplot(data = rates, aes(x = date, y = rate, linetype = type))+
  geom_line(lwd = 1)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, hjust = 0, angle = 45),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("interest rate (%)") +
  labs(title = "Interest rates in the United Kingdom, 1995-2023",
       caption = "Source: Bank of England. Series: IUMABEDR, IUMODTL and IUMTLMV.")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8.7, fig.width = 14, fig.align="center"}
p1
```





---
`r chunk_reveal("chunk_11")`

```{r chunk_11, include = FALSE}
read_csv("../data/IUMABEDR  Bank of England  Database.csv", 
                         col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Policy rate')



read_csv("../data/IUMTLMV  Bank of England  Database.csv", 
                        col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. mortgage rate')


mortgage_rate <- read_csv("../data/IUMODTL  Bank of England  Database.csv", 
                          col_names = FALSE, skip = 1) %>% 
  rename(rate = X2,
         date = X1) %>% 
  mutate(date = as.Date(date, format = "%d %b %y")) %>% 
  mutate(type = 'Av. overdraft rate')

rates <- rbind(policy_rate, mortgage_rate, overdraft_rate)
```




---
`r chunk_reveal("chunk_12")`

```{r chunk_12, include = FALSE}
ggplot(data = rates, aes(x = date, y = rate, linetype = type))+
  geom_line(lwd = 1)+
    scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("interest rate (%)") +
  labs(title = "Interest rates in the United Kingdom, 1995-2023",
       caption = "Source: Bank of England. Series: IUMABEDR, IUMODTL and IUMTLMV.")+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, hjust = 0, angle = 45),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())
```





---

## Factors affecting the interest rate

There are many factors shaping the interest rates in the economy: 

- The cost of obtaining money (i.e. the base rate and the interbank interest rate).

- The risk of the loan.

- The time span of the loan.

- Whether the interest rate is fixed or flexible.







---

## The real interest rate

Data on interest rates are typically stated in nominal terms

- If I lend you money for a 2% annual interest rate and the inflation is 3% I lose.

- I can buy more for the 100£ I give you today, than for the 102£ you give me in one year.


The link between the real ($r$) and nominal interest $i$ rates are is the following: 

$$(1+i) = (1+r)(1+\pi) = 1 + r + \pi + \pi r$$

The *Fisher equation*: $r = i - \pi$

- The indentity is based on the assumption that $\pi r\approx 0$.







---
class: inverse, center, middle

# Exchange Rates 

---



## Exchange rates

The rate at which we are willing to trade on currency to another currency

“An exchange rate of 1.39 US Dollar to the British Pound.”

“An exchange rate of 1.39 US Dollar to buy one British Pound.”




---





## Exchange rates: ups and downs

What determines the exchange rate?

- Financial institutions trade currencies on the foreign exchange market (forex).

... and then buy and sell them to us (we pay a premium)

**Appreciates**: the currency increases in value – ex. 1£ from costing 1.39$ to 1.45$

**Depreciates**: the currency loses in value – ex. 1£ from costing 1.39$ to 1.28$







---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
ex_rate <- read_table2("../data/Exchange_Rate_Report.tsv", 
                                    skip = 2) %>% 
  rename(rate = `U.K.`) %>% 
  mutate(date = as.Date(Date, format = "%d-%b-%Y")) %>% 
  select(date, rate) %>% 
  mutate(g = (rate - lag(rate, n = 1))/lag(rate, n = 1))


p1 <- ggplot(data = ex_rate, aes(x = date, y = rate))+
  geom_line()+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, angle=45, hjust=1),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("USD/GBP") +
  labs(title = "GBP against the US dollar, 1995-2023",
       caption = "Source: International Monetary Fund. Series: Representative rates.")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8.7, fig.width = 14, fig.align="center"}
p1
```









---

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE,fig.align="center", include=FALSE}
ex_rate <- read_table2("../data/Exchange_Rate_Report.tsv", 
                                    skip = 2) %>% 
  rename(rate = `U.K.`) %>% 
  mutate(date = as.Date(Date, format = "%d-%b-%Y")) %>% 
  select(date, rate) %>% 
  mutate(g = (rate - lag(rate, n = 1))/lag(rate, n = 1))


p1 <- ggplot(data = ex_rate, aes(x = date, y = g))+
  geom_line(lwd = 0.5)+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, angle=45, hjust=1),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("USD/GBP") +
  labs(title = "GBP against the US dollar, 1995-2023",
       caption = "Source: International Monetary Fund. Series: Representative rates.")

```

```{r dev.args = list(bg = 'transparent'), fig.align="center", echo=FALSE, message=FALSE, warning=FALSE, fig_caption= FALSE, fig.height = 8.7, fig.width = 14, fig.align="center"}
p1
```






---
`r chunk_reveal("chunk_21")`

```{r chunk_21, include = FALSE}
read_table2("../data/Exchange_Rate_Report.tsv", 
                                    skip = 2) %>% 
  rename(rate = `U.K.`) %>% 
  mutate(date = as.Date(Date, format = "%d-%b-%Y")) %>% 
  select(date, rate) %>% 
  mutate(g = (rate - lag(rate, n = 1))/lag(rate, n = 1))
```




---
`r chunk_reveal("chunk_22")`

```{r chunk_22, include = FALSE}
ggplot(data = ex_rate, aes(x = date, y = g))+
  geom_line(lwd = 0.5)+
    scale_x_date(date_breaks = "24 month", date_labels = "%Y")+
  xlab("") +
  ylab("USD/GBP") +
  labs(title = "GBP against the US dollar, 1995-2023",
       caption = "Source: International Monetary Fund. Series: Representative rates.")+
  theme(plot.title = element_text(size=22),
        plot.subtitle = element_text(size=16),
        legend.title = element_blank(),
        legend.text=element_text(size=14),
        legend.position="bottom",
        plot.caption = element_text(size=12, hjust = 0),
        axis.title = element_text(size = 18),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 14, angle=45, hjust=1),
        axis.text.y=element_text(size = 8.75),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())
```


---
## Some date formats in R


%d = Day number of month (5, 17, 28, etc.)

%a = Abbreviated weekday (Mon, Tue, Wed, etc.)

%A = Full weekday (Monday, Tuesday, etc.) %w = Weekday number (0-6, Sunday is 0)

%u = Weekday number (1-7, Monday is 1)

%W = Week number (00-53, Monday is week start)

%U = Week number (01-53, Sunday is week start)

%m = Month number (e.g. 01, 02, 03, 04)

%b = Abbreviated month (Jan, Feb, etc.)

%B = Full month (January, February, etc.)

%y = 2-digit year (e.g. 89)

%Y = 4-digit year (e.g. 1989)







---



## After this lecture you should be able to

Explain the 3 different types of interest rates.

- Bank rate (Policy rate): The central bank’s rate

- Interbank rate: the rate between banks

- Household rate: the rate we are paying.


Explain why interest vary 

- Information and security

- Flexibility

- Duration


Describe exchange rates

- Expressed in how much it costs to buy 1 currency unit.

- Appreciation/depcreciation: changes relative to other currencies.








---

class: inverse, center, middle

# See you in the labs! 

Questions? $\texttt{sebastian.ellingsen@bristol.ac.uk}$.







